---
title: "COVID-19 Death Cases Analysis by Date, State, Sex and Age Groups"
author: "Chen Chen"
output: 
    html_document:
        toc: TRUE
        toc_float: TRUE
---


This is PM566 final project website. 
Click to [Download the Report](https://github.com/cchen941/PM566-finalproject/blob/main/final_report/FinalReport.pdf)


Click to [View the Dataset](https://data.cdc.gov/NCHS/Provisional-COVID-19-Deaths-by-Sex-and-Age/9bhg-hcku)

<br>


```{r setup, message=FALSE, echo=FALSE, warning=FALSE}
library(data.table)
library(lubridate)
library(tidyverse)
library(dplyr)
library(plotly)
library(DT)
library(knitr)
data <- data.table::fread("data/Provisional_COVID-19_Deaths_by_Sex_and_Age.csv")

# Initialize code chunk options
opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  eval=TRUE,
  echo = TRUE,
  cache = FALSE,
  fig.width = 7,
  fig.align = 'center',
  fig.asp = 0.618,
  out.width = "700px")
```

```{css, echo = FALSE}
.code-r { /* Code block */
  font-size: 15px;
}

.code-r-small { /* Code block */
  font-s
  ize: 10px;
}
```


```{r echo=FALSE, warning= FALSE}
data_clean <- select(data, 3:11)
data_clean <- filter(data_clean, Group == "By Month", State != "United States", Sex != "All Sexes", `Age Group` != "All Ages")
data_clean <- data_clean[!is.na(`COVID-19 Deaths`)]
data_clean <- data_clean[!is.na(`Total Deaths`)]
```


## Brief Description
As we know, most people infected with COVID-19 will experience mild to moderate respiratory illness and recover without requiring special treatment. However, some will become seriously ill or die at any age. The data set of this project is from CDC (Centers for Disease Control and Prevention) and talks about provisional COVID-19 deaths in United States by date, state, sex and age groups.


The objective of this project is to explore the answer to questions below: 

Do COVID-19 death cases decrease by date?

Do COVID-19 death cases vary by state, sex, age groups?

<br>

## Interacitve Tables and Visualizations
### Line Plot of Covid-19 Deaths by Date

```{r echo=FALSE, warning=FALSE, class.source="code-r-small"}
data_p1 <-filter(data, Group == "By Month", State != "United States", Sex == "All Sexes", `Age Group` == "All Ages")
data_p1$`End Date` <- as.Date(data_p1$`End Date`, format="%m/%d/%Y")
p1 <- ggplot(data_p1, aes(x= `End Date`, y= `COVID-19 Deaths`, color = State)) + geom_line()+ geom_point(size= 0.5, alpha =0.5)+ xlab("Date")
p1 <- ggplotly(p1)
p1
```


The first chart is the line plot of Covid-19 death cases by date. The range of the date is from January 2020 to November 2022. The different color here in the plot represents the different state. By this, we can find out distribution by date among states as well. From the chart, We can find that from 2020 to 2022, the overall number of deaths shows a trend of increasing first and then decreasing. The data peaks in early 2021, and the recent death data (end of 2022) is much smaller than the data when the virus just come out to spread (2020). It indicates that the death rate from Covid-19 has been greatly reduced. From the perspective of states, the area with the largest number of deaths in 2020 in New York City. It is noticed that blue lines are at higher levels, referring to the vicinity of New York State. After entering 2021, California's data began to rise rapidly and reached the highest peak. At the end of 2021, there is a decline in CA, and the peaks become Florida and Texas.

<br>

### Bar Chart of Covid-19 Deaths by Age Groups

```{r echo=FALSE, warning=FALSE, class.source="code-r-small"}
p2<-data_clean%>%
ggplot(aes(x=`COVID-19 Deaths`, y=fct_reorder(`Age Group`, `COVID-19 Deaths`),fill=Sex))+ geom_col(position= "dodge")+ ylab("Age Groups")
p2 <- ggplotly(p2)
p2
```


From the chart,the number of death cases shows an apparent increasing trend with increasing age. It is worth mentioning that the bars of different colors in the figure refer to different genders: blue represents males, and red represents females. From the chart, except for the two age groups 55-64 and over 85, most of the other groups have more male deaths than females. This is a gender-related result that can be reflected in this graph.

<br>

### Histogram of Covid-19 Deaths Counts under 250
```{r echo=FALSE, warning=FALSE, class.source="code-r-small"}
data_p3 <- data_clean[`COVID-19 Deaths`<250]
data_p3 <- data_p3[`COVID-19 Deaths`>0]
p3<- data_p3%>% ggplot(aes(x=`COVID-19 Deaths`, fill=Sex))+ geom_histogram(position= "dodge",binwidth = 5)
p3 <- ggplotly(p3)
p3
```
For sex groups analysis, the next figure is one histogram of death cases colored by gender. The figure only shows the data with the count of death cases below 250. Because after observing the dataset, most of the observations are below 250. In this case, the distribution of gender classification can be indicated more intuitively. It can be clearly seen from the histogram that most of the observations for death cases under 250 have more males than females. Combining the above visualization results, we can conclude that in the three-year Covid-19 death data from 2020 to 2022, there are more males than females.

<br>


### Scatter Plot of Covid-19 Deaths by Total Deaths
```{r echo=FALSE, warning=FALSE, class.source="code-r-small"}
data_clean%>%
  plot_ly(x= ~`COVID-19 Deaths`, y= ~`Total Deaths`, 
          type = 'scatter',
          mode = 'markers',
          color = ~State,
          marker = list(sizemode = 'sex', opacity= 0.5))
```
This plot shows the relationship between the number of Covid-19 deaths and the total number of deaths. It is not difficult to find that these two variables have a certain degree of correlation. With the increase in the total deaths, number of deaths from Covid-19 is close to linear growth. From the perspective of states, this interactive plot helps show those outlier data points.

<br>

## Conclusion and Summary
During these three years, the number of Covid-19 deaths has fluctuated, showing an increasing trend until the beginning of 2021 and starting to decline significantly thereafter; For the data of each state, California, Texas, and Florida have the top three most total Covid-19 death data. New York City, California, and Florida have all reached the highest level among states at different times; The Covid-19 death cases increase by age group, with higher age groups having more deaths; In terms of gender, most of the total number and proportion of Covid-19 deaths in the male group is higher the female group.

